A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-10 podc_paper/correctness.tex
cutting to save space

\section{Proof of Correctness}

@@

Now we show $v.\fld{blocks}[v.\head]$ is either the last non-\nl\ block or the first \nl\ block in node $v$.

\begin{invariant}\label{lem::headPosition} 
For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
If $v\neq \var{root}$,  $v.\fld{blocks}[i].super \neq \nl$ for $0<i<v.head$.
\end{invariant}

\begin{proof}
Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.

Assume the claim holds before a change to $v.\fld{blocks}$, which can be made only
by a successful \op{CAS} at line \ref{cas}.
The \op{CAS} changes $v.\fld{blocks}[h]$ from \nl\ to a non-\nl\ value.
Since $v.\fld{blocks}[h]$ is \nl\ before the CAS, $v.\head \leq h$ by the hypothesis.
Since $h$ was read from $v.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
$v.\head \geq h$ by \Cref{lem::headPosition}.
So, $h=v.\head$ and a change to $v.\fld{blocks}[v.\head]$ preserves the invariant.

Now, assume the claim holds before a change to $v.\head$, which can only be an increment from $h$ to $h+1$
by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
For the first two claims, it suffices to show that $v.\fld{blocks}[head] \neq \nl$.
\nf{Advance} is called either at line \ref{helpAdvance} 
after testing that $v.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$.
For the third claim, observe that prior to incrementing $v.\head$ at line \ref{incrementHead},
the \op{CAS} at line \ref{setSuper1} ensures that $v.\fld{blocks}[i].super\neq \nl$.
\end{proof}

@@

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $v.\fld{blocks}[b] \neq \nl$ then 
 $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
 $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
\end{lemma}
\begin{proof}
Let $B$ be the block in $v.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $v.\head$ at line \ref{readHead}.
At that time, $v.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
Thus, the \op{CreateBlock}($v,b-1$) that created $B'$ terminated before the \op{CreateBlock}($v,b$) that
created $B$ started.
It follows from \Cref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($v,b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($v,b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
%of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

@@

\here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
\begin{corollary}\label{lem::noDuplicates}
For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
\end{corollary}
\begin{proof}
The operations contained in a block $B$ are those that appear in subblocks of $B$ in the leaves of the tree.
Since each process puts each of its operations in only one block of its own leaf, an operation 
cannot be in two different leaf blocks. 
By \Cref{lem::subblocksDistinct}, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ have no subblocks in common, so the claim follows.
\end{proof}

@@

The following shows that the values stored in \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields are accurate.
\here{If space is tight, could easily move next proof to appendix and just say it follows easily from the definition of subblocks and the values stored in the fields on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq}, \ref{createSumDeq}.}

\begin{invariant}\label{lem::sum}
If $B$ is a block stored in $v.\fld{blocks}[i]$,
$B.\fld{sum\sub{enq}} = | E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i]) |$ and
$B.\fld{sum\sub{deq}} = | D(v.\fld{blocks}[0])\cdots D(v.\fld{blocks}[i]) |$.
\end{invariant}
\begin{proof}
Initially, each \fld{blocks} array only contains an empty block $B_0$ in location 0.
By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
Moreover, $B_0.\fld{sum\sub{enq}} = B_0.\fld{sum\sub{deq}} = 0$, so the claim is true.

We show that each installation of a block $B$ into some location $v.\fld{blocks}[i]$ preserves the claim,
assuming the claim holds before this installation.  We consider two cases.

If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.

Now suppose $v$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
subblocks of $v.\fld{blocks}[1..i]$ are $v.\fld{left.blocks}[1..B.\eleft]$ 
and $v.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i])$ are those in
$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eleft])$ and
$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is $v.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

\begin{corollary}\label{blockNotEmpty}
If a block $B$ is in $v.\fld{blocks}[i]$ where $i>0$, then $E(B)$ and $D(B)$ are not both empty.
\end{corollary}
\begin{proof}
The \op{Refresh} that installed $B$ got $B$ as the response to its call to \op{CreateBlock} on line \ref{invokeCreateBlock}.
Thus, at line \ref{testEmpty} $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
By \Cref{lem::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
so these sequences cannot both be empty.
\end{proof}

@@

Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
We say that node $v$ \emph{contains} an operation if some block in $v.\fld{blocks}$ contains the operation.
%\here{move this defn earlier and just recall it here?}
Once a block is added to a node, it remains there forever.  Thus, if $v$ contains an operation at some time, it contains the operation at all later times too.

\begin{lemma}\label{successfulRefresh}
Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
when $R$ executed line~\ref{readHead}.
\end{lemma}
\begin{proof}
Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{readHead}.
Let $i$ be the index such that the block $B=v.\fld{left.blocks}[i]$ contains $op$.
By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
$v.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
$v.\fld{left.head} > i$.
Then, $R$ calls \op{CreateBlock}($v,h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
\op{CreateBlock} reads a value greater than $i$ from $v.\fld{left.head}$ at line \ref{createEndLeft}.
Thus, $new.\eleft \geq i$.  We consider two cases.

Suppose $R$'s call to \op{CreateBlock} returns the new block $B'$ and $R$'s \op{CAS} at line \ref{cas} 
installs $B'$ in $v.\fld{blocks}$.
Then, $B$ is a subblock of some block in $v$, since  $B'.\eleft$ is greater than or equal to $B$'s index
in $v.\fld{left.blocks}$.
Hence $v$ contains $op$, as required.

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
\end{proof}

We now show that if a process fails its \op{Refresh}($v$) twice, some other process must have succeeded.

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
All operations contained $v$'s children when $R_1$ begins
are contained in $v$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

@@

We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  This is proved by showing that the \fld{super} field of a block $B$ in node $v$ is read from the 
\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
On one hand, $B.super$ is written before $B_s$ is installed:
\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
$B.\fld{super}$ is written after $B$ is installed, and \Cref{successfulRefresh} ensures that $B$ is propagated to the
parent soon after.

\begin{restatable}{lemma}{superRelationRes}
\label{superRelation}
Let $B=v.\var{blocks}[b]$.
  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{restatable}

@@

\begin{restatable}{lemma}{indexDequeueRes}
\label{lem::indexDequeue}
If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($v, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{restatable}

\begin{restatable}{lemma}{getEnqRes}
\label{lem::get}
If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
\end{restatable}

>>>>

\section{Proof of Correctness}
\label{sec::correctness}

@@

Now we show $\var{v}.\fld{blocks}[\var{v}.\head]$ is either the last non-\nl\ block or the first \nl\ block in node $\var{v}$.

\begin{invariant}\label{lem::headPosition} 
For $0 \leq i < \var{v}.\head$, $\var{v}.\fld{blocks}[i]\neq\nl$.  For $i>\var{v}.\head$, $\var{v}.\fld{blocks}[i]=\nl$.
If $\var{v}\neq \var{root}$,  $\var{v}.\fld{blocks}[i].super \neq \nl$ for $0<i<\var{v}.\head$.
\end{invariant}

\begin{proof}
Initially, $\var{v}.\head=1$, $\var{v}.\fld{blocks}[0]\neq\nl$  and $\var{v}.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.

Assume the claim holds before a change to $\var{v}.\fld{blocks}$, which can be made only
by a successful \op{CAS} at line \ref{cas}.
The \op{CAS} changes $\var{v}.\fld{blocks}[h]$ from \nl\ to a non-\nl\ value.
Since $\var{v}.\fld{blocks}[h]$ is \nl\ before the CAS, $\var{v}.\head \leq h$ by the hypothesis.
Since $h$ was read from $\var{v}.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
$\var{v}.\head \geq h$ by \Cref{lem::headPosition}.
So, $h=\var{v}.\head$ and a change to $\var{v}.\fld{blocks}[\var{v}.\head]$ preserves the invariant.

Now, assume the claim holds before a change to $\var{v}.\head$, which can only be an increment from $h$ to $h+1$
by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
For the first two claims, it suffices to show that $\var{v}.\fld{blocks}[head] \neq \nl$.
\nf{Advance} is called either at line \ref{helpAdvance} 
after testing that $\var{v}.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $\var{v}.\fld{blocks}[h]\neq\nl$.
For the third claim, observe that prior to incrementing $\var{v}.\head$ at line \ref{incrementHead},
the \op{CAS} at line \ref{setSuper1} ensures that $\var{v}.\fld{blocks}[i].super\neq \nl$.
\end{proof}

@@

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$ then 
 $\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{right}}$.
\end{lemma}
\begin{proof}
Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $\var{v}.\head$ at line \ref{readHead}.
At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
created $B$ started.
It follows from \Cref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
%of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

@@

\here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
\begin{corollary}\label{lem::noDuplicates}
For  $i\neq j$, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ cannot both contain the same operation.
\end{corollary}
\begin{proof}
A block $B$ contains the operations in $B$'s subblocks in leaves of the tree.
An operation by process $P$ appears in one block of $P$'s leaf, so
an operation 
cannot be in two different leaf blocks. 
By \Cref{lem::subblocksDistinct}, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ have no common subblocks, so the claim follows.
\end{proof}

@@

The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
on line \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
from the definition of subblocks.  See Appendix 
\ref{app::tracingDetails} for a detailed proof.

\begin{restatable}{invariant}{sumRes}
\label{lem::sum}
If $B$ is a block stored in $\var{v}.\fld{blocks}[i]$,
$B.\fld{sum\sub{enq}} = | E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i]) |$ and
$B.\fld{sum\sub{deq}} = | D(\var{v}.\fld{blocks}[0])\cdots D(\var{v}.\fld{blocks}[i]) |$.
\end{restatable}

@@

\begin{corollary}\label{blockNotEmpty}
If a block $B$ is in $\var{v}.\fld{blocks}[i]$ where $i>0$, then $E(B)$ and $D(B)$ are not both empty.
\end{corollary}
\begin{proof}
The \op{Refresh} that installed $B$ got $B$ as the response to its call to \op{CreateBlock} on line \ref{invokeCreateBlock}.
Thus, at line \ref{testEmpty} $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
By \Cref{lem::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
so these sequences cannot both be empty.
\end{proof}

@@

Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
%\here{move this defn earlier and just recall it here?}
Since blocks are permanently added to nodes, if $\var{v}$ contains $op$ at some time, $v$ contains $op$ at all later times too.

\begin{lemma}\label{successfulRefresh}
Let $R$ be a call to \op{Refresh}($\var{v}$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
In the configuration after that CAS (or termination, respectively), $\var{v}$ contains all operations that $\var{v}$'s children contained 
when $R$ executed line~\ref{readHead}.
\end{lemma}
\begin{proof}
Suppose $\var{v}$'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{readHead}.
Let $i$ be the index such that the block $B=\var{v}.\fld{left.blocks}[i]$ contains $op$.
By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
$\var{v}.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
$\var{v}.\fld{left.head} > i$.
Then, $R$ calls \op{CreateBlock}($\var{v},h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
\op{CreateBlock} reads a value greater than $i$ from $\var{v}.\fld{left.head}$ at line \ref{createEndLeft}.
Thus, $new.\eleft \geq i$.  We consider two cases.

Suppose $R$'s call to \op{CreateBlock} returns the new block $B'$ and $R$'s \op{CAS} at line \ref{cas} 
installs $B'$ in $\var{v}.\fld{blocks}$.
Then, $B$ is a subblock of some block in $\var{v}$, since  $B'.\eleft$ is greater than or equal to $B$'s index
in $\var{v}.\fld{left.blocks}$.
Hence $\var{v}$ contains $op$, as required.

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

We now show that if a process fails its \op{Refresh}($\var{v}$) twice, some other process must have succeeded.

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($\var{v}$) by the same process.
All operations contained $\var{v}$'s children when $R_1$ begins
are contained in $\var{v}$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $\var{v}.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $\var{v}.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $\var{v}$'s children when $R_1$ begins
are also contained in $\var{v}$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

@@

We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from the 
\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
On one hand, $B.super$ is written before $B_s$ is installed:
\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
$B.\fld{super}$ is written after $B$ is installed, and \Cref{successfulRefresh} ensures that $B$ is propagated to the
parent soon after.

\begin{restatable}{lemma}{superRelationRes}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
  If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{restatable}

@@

\begin{restatable}{lemma}{indexDequeueRes}
\label{lem::indexDequeue}
If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{restatable}

\begin{restatable}{lemma}{getEnqRes}
\label{lem::get}
If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
\end{restatable}
